Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-03-10 paper.qmd
Removed unecessary code

hex_grid_coloured_two_curvy1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts, 
    aes(x = x, y = y,
      group = hex_poly_id, 
      fill = std_counts), color = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
             aes(x = tSNE1, y = tSNE2),
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "E") +
  interior_annotation("b", sc_ltr_pos, cex = 1.5) + 
  xlim(sc_xlims) + ylim(sc_ylims)

@@

# trimesh_removed_two_curvy <- vis_rmlg_mesh(
#   distance_edges = distance_two_curvy, 
#   benchmark_value = benchmark_two_curvy, 
#   tr_coord_df = tr_from_to_df_two_curvy, 
#   distance_col = "distance") +
#   xlim(sc_xlims) + ylim(sc_ylims) +
#   interior_annotation("d", sc_ltr_pos)

@@

<!--langevitour with true model for S-curve--> 
<!-- ::: {#fig-two_curvy-true-sc layout-ncol="3" fig-pos="H"} -->

<!-- ![](figures/two_curvy/sc_true_1.png){width="150" fig-align="center"} -->

<!-- ![](figures/two_curvy/sc_true_2.png){width="150" fig-align="center"} -->

<!-- ![](figures/two_curvy/sc_true_3.png){width="150" fig-align="center"} -->

<!-- Three views of the true model (grey points and lines) in $2\text{-}D$ projections from $7\text{-}D$, for the S-curve data (purple points). The data is spread along the S shape, and does not vary much from this curve. (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/I5GL23vLiw0>).  -->
<!-- ::: -->

<!-- XXX The points on the "theoretical model" should be equidistant along the S. They are still to bunched in some spots and too far apart in others. -->

@@

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_30_min_dist_0.3.rds")
pbmc_scaled_obj <- gen_scaled_data(
  data = umap_pbmc)
umap_pbmc_scaled <- pbmc_scaled_obj$scaled_nldr

# umap_pbmc <- umap_pbmc_scaled |>
#   ggplot(aes(x = UMAP1,
#              y = UMAP2)) +
#   geom_point(alpha=0.3, color = "#000000") 
```

@@

# trimesh_removed_pbmc <- vis_rmlg_mesh(
#   distance_edges = distance_pbmc,
#   benchmark_value = benchmark_pbmc,
#   tr_coord_df = tr_from_to_df_pbmc,
#   distance_col = "distance") +
#   geom_point(
#     data = umap_pbmc_scaled,
#     aes(
#       x = UMAP1,
#       y = UMAP2
#     ),
#     alpha = 0.3,
#     size = 0.1,
#     color = "#000000"
#   )

trimesh_removed_pbmc <- ggplot() + 
  geom_segment(data = tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  geom_point(
    data = umap_pbmc_scaled,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    alpha = 0.2,
    color = '#a65628'
  )  +
  interior_annotation("a1", cex = 2) +
  theme(
    aspect.ratio = 1
  )

# ## Compute error
# error_df <- augment(
#   df_bin_centroids = df_bin_centroids_pbmc,
#   df_bin = df_bin_pbmc,
#   training_data = training_data_pbmc,
#   newdata = NULL,
#   type_NLDR = "UMAP",
#   col_start = "PC_")

# ## Categorize error
# 
# error_df <- error_df |>
#   mutate(type = case_when(
#     row_wise_abs_error <= 5 ~ "error 0-5",
#     row_wise_abs_error <= 10 ~ "error 5-10",
#     row_wise_abs_error <= 15 ~ "error 10-15",
#     row_wise_abs_error <= 20 ~ "error 15-20",
#     .default = "error greter than 20"
#   )) |>
#   mutate(type = factor(type, levels = c(
#     "error 0-5", "error 5-10", "error 10-15", "error 15-20", "error greter than 20")))
# 
# ## To join embedding
# error_df <- error_df |>
#   bind_cols(umap_pbmc_scaled |>
#               select(-ID))
# 
# error_plot_pbmc <- error_df |>
#   ggplot(aes(x = UMAP1,
#              y = UMAP2,
#              color = type,
#              group = ID)) +
#   geom_point(alpha=0.5,
#              size = 0.1) +
#   interior_annotation("b", sc_ltr_pos)

@@

# projection_filtered <- projection
# projection_filtered[apply(projection_filtered, 1, function(row) all(row < 0)), ] <- 0.75

#rows_to_keep <- c(1, 2, 4, 5, 6)

#projection_filtered <- projection[rows_to_keep, ]

@@

# pbmc_proj_umap_model2 <- ggplot() +
#   geom_segment(
#     data = model_df, 
#     aes(
#       x = proj1_from, 
#       y = proj2_from, 
#       xend = proj1_to, 
#       yend = proj2_to), 
#     color = "#000000") +
#   geom_segment(
#     data=axes, 
#     aes(x=x1, y=y1, xend=x2, yend=y2), 
#     colour="grey70") +
#   geom_text(
#     data=axes,
#     aes(x=x2, y=y2), 
#     label=rownames(axes),
#     colour="grey50",
#     size = 3) +
#   geom_path(
#     data=circle, 
#     aes(x=c1, y=c2), colour="grey70") +
#   coord_fixed() +
#   xlim(c(-0.65, 0.65)) +
#   ylim(c(-0.65, 0.65)) +
#   interior_annotation("a3", cex = 1.5)

@@

```{r}
#| label: error-comp-pbmc
# read the png file from device

#pbmc_layouts <- readPNG("figures/pbmc3k/pbmc_nldr_layouts.png", native = TRUE)

error_plot_pbmc <- ggplot(error_pbmc, 
                          aes(x = a1, 
                              y = MSE, 
                              colour = method)) + 
  geom_point(size = 0.8) +
  geom_line(linewidth = 0.3) + 
  # geom_vline(xintercept = 15, linetype="solid", 
  #            color = "black", linewidth=0.8, alpha = 0.5) +
  scale_x_continuous(breaks = sort(unique(error_pbmc$a1))[c(1, 5, 9, 13, 17, 21, 26)]) +
  scale_color_manual(values=c('#e41a1c','#377eb8','#4daf4a','#ff7f00',
                              '#984ea3','#999999','#a65628','#f781bf')) +
  scale_y_log10() +
  ylab("log(MSE)") +
  xlab(expression(paste("binwidth (", a[1], ")"))) +
  theme_minimal() +
  theme(panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        legend.position = "none",
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7))

@@

distance_df_small_edges_pbmc <- distance_pbmc |>
  filter(distance < benchmark_pbmc) #|>
  #filter(!(row_number() %in% c(6, 12, 14, 16, 17, 19, 23, 29, 30, 143, 144, 146, 147, 157)))

@@

# tr_from_to_df_pbmc <- tr_from_to_df_pbmc |>
#   filter(!(row_number() %in% c(155)))

@@

# trimesh_removed_pbmc <- vis_rmlg_mesh(
#   distance_edges = distance_pbmc,
#   benchmark_value = benchmark_pbmc,
#   tr_coord_df = tr_from_to_df_pbmc,
#   distance_col = "distance") +
#   geom_point(
#     data = tsne_pbmc_scaled,
#     aes(
#       x = tSNE1,
#       y = tSNE2
#     ),
#     alpha = 0.2,
#     color = "#000000"
#   ) 
#   #xlim(sc_xlims) + ylim(sc_ylims) +
#   #interior_annotation("a", sc_ltr_pos)

@@

## Removed long edge in high-d
# distance_df_small_edges_pbmc <- distance_df_small_edges_pbmc |>
#   filter(!(row_number() %in% c(155)))

>>>>

hex_grid_coloured_two_curvy1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts, 
    aes(x = x, y = y,
      group = hex_poly_id, 
      fill = std_counts), color = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
             aes(x = tSNE1, y = tSNE2),
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "E") +
  interior_annotation("b", sc_ltr_pos, cex = 1.5) + 
  xlim(sc_xlims) + ylim(sc_ylims) 
```

@@

@@

@@

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_30_min_dist_0.3.rds")
pbmc_scaled_obj <- gen_scaled_data(
  data = umap_pbmc)
umap_pbmc_scaled <- pbmc_scaled_obj$scaled_nldr
```

@@

trimesh_removed_pbmc <- ggplot() + 
  geom_segment(data = tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  geom_point(
    data = umap_pbmc_scaled,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    alpha = 0.2,
    color = '#a65628'
  )  +
  interior_annotation("a1", cex = 2) +
  theme(
    aspect.ratio = 1
  )
```

@@

@@

@@

```{r}
#| label: error-comp-pbmc

error_plot_pbmc <- ggplot(error_pbmc, 
                          aes(x = a1, 
                              y = MSE, 
                              colour = method)) + 
  geom_point(size = 0.8) +
  geom_line(linewidth = 0.3) + 
  scale_x_continuous(breaks = sort(unique(error_pbmc$a1))[c(1, 5, 9, 13, 17, 21, 26)]) +
  scale_color_manual(values=c('#e41a1c','#377eb8','#4daf4a','#ff7f00',
                              '#984ea3','#999999','#a65628','#f781bf')) +
  scale_y_log10() +
  ylab("log(MSE)") +
  xlab(expression(paste("binwidth (", a[1], ")"))) +
  theme_minimal() +
  theme(panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        legend.position = "none",
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7))

@@

@@

@@

@@
